### 2022-09-13 AISTATS2023/paper.tex
reached v1 of method description

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML system with the capability to
explain their predictions, apart from predicting with accuracy. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local ones into a single interpretable outcome, usually a
number or a plot. If a user wants to get a rough overview about which
features are significant (feature importance) or whether a particular
feature has a positive or negative effect on the output (feature
effect), they should opt for a global explainability technique. On
ther other hand, aggregating the individual explanations for producing
a concise global one is vulnerable to misinterpretations; Under strong
feature interactions, the global explanation may obfuscate
heterogeneous effects~\citep{Herbinger2022repid} that exist under the
hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
of global explainability methods. The objective of FE is to the
isolate and visulalize the impact of a single feature on the
output.~\footnote{FE methods also isolate the effect of a pair of
  features to the output. Combinations of more than two features are
  not usual, because they encounter, among others, visualization
  difficulties.} FE methods suffer from aggregation bias because,
often, the rationale behind the average effect might be unclear. For
example, a feature with zero average effect may indicate that the
feature has no effect on the output or, contrarily, it has a highly
positive effect in some cases and a highly negative in others. There
are three widely-used FE methods; Partial Dependence Plots
(PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML
problems. Therefore, ALE has been established as the state-of-the-art
FE method.

However, ALE faces two crucial drawbacks. First, it does not provide a
way to inform the user about potential heterogeneous effects that are
hidden behind the average effect. In contrast, in the case of PDP, the
heterogeneous effects can be spotted by exploring the Individual
Conditional Expectations (ICE)\citep{goldstein2015peeking}. Second,
ALE requires an additional step where the axis of the feature of
interest is split in \(K\) fixed-size non-overlapping intervals. So
far, the user is asked to provide a value for the parameter \(K\)
blindly, i.e without an indication about whether a big or a small
value could be more appropriate, which often leads to unstable
explanations. For convenience, in the rest of the paper we will refer
to this step using the term bin-splitting problem.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the global explanation. The uncertainty
of the global explanation expresses how certain we are that the global
(averaged) explanation is valid if applied to an instance drawn at
random. The probabilistic extension completes ALE, as ICE plots
complement PDP, for revealing the heterogeneous effects. Furthermore,
we transform the bin-splitting step into a data-driven clustering
problem, i.e., we search for the optimal splitting given available
instances of the training set. We, also, present a
computationally-grounded algorithm for finding the optimal
solution.

\paragraph{Contributions.} The contributions of this paper are the following:

\begin{itemize}
\item We introduce NAME, an extension of ALE that quantifies the
  uncertainty of the global explanation, i.e.~the level of
  heterogeneous effects hidden behind the global explanation.
\item We present an algorithm that automatically computes the optimal
  bin-splitting
\item We formally prove that our method finds the optimal grouping of
  samples, minimizing the added uncertainty over the unavoidable
  heterogeneity.
\item We provide empirical evaluation of the method in artificial and
  real datasets.
\end{itemize}

The implementation of our method and the code for reproducing all the
experiments is provided in the submission and will become publicly
available upon acceptance.

@@

\paragraph{Feature Effect Methods.} The three well-known feature
effect methods are: PDP, MP and ALE. PDP formulates the FE
of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects:

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
ALE has specific advantages which gain particular value in cases of
corelated input features. In these cases, PDP integrates over
unrealistic instances, due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore the only trustable
FE method in cases of correlated features.

@@

FE methods answer the question \textit{what is expected to happen to
  the output (expected effect), if the value of a specific feature is
  increased/decreased}. Having an answer to the question above, it
comes naturally to also wonder \textit{how certain we are about the
  expected change (uncertainty)}. For this reason, the quantification
of the uncertainty along with the expected effect has attracted a lot
of interest. The level of uncertainty is mostly quantified by
measuring the existence of heterogeneous effects, i.e. whether there
are local explanations that deviate from the expected global
effect. ICE and d-ICE plots provide a visual understanding of the
heterogeneous effects on top of PDPs. Another approach targets on
grouping the heterogeneous effects, e.g., allocating ICE plots in
homogeneous clusters, by dividing the input
space.\citep{molnar2020model} Some other approaches, like H-Statistic,
Greenwel, move a step behind and try to quantify the level of
interaction between the input features, a possible cause of
heterogeneous effects. In this case, the interpretation is indirect,
since a strong interaction index is only an indicator of heterogeneous
effects. The aforementioned approaches face two pathogenies; They
either do not quantify the uncertainty of the FE directly or they are
based on PDPs, and, therefore, they are subject to the failure modes
of PDPs in cases of correlated features. To the best of our knowledge,
no method so far targets on quantifying the uncertainty of ALE.

@@

In real ML scenarios, the expected FE and the uncertainty are
estimated from the limited instances of the training set.
\citep{apley2020visualizing} proposed estimating the local effects in
each bin by evaluating the black box-funtion at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE) that
exploits differentiaton for computing the local effects on the instances
training-set, instead of the bin limits:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost. However, none of the approximations above deals
with the crucial problem of bin-spliting. As indicated
by~\citep{molnar2022}, in ALE the effects are computed per interval
(region) and the interpretation of the effect can only be
local.

@@

In Section~\ref{sec:NAME-definition} we define the component for the
uncertainty quantification. In
Section~\ref{sec:interval-based-estimation}, we show how to estimate
the average effect and the uncertainty from the limited samples of the
training set and we make an important proof about the aggregated
variance defined over a bin. In Section~\ref{sec:interval-spliting},
we define and solve the problem of optimal bin-splitting. Finally, in
Section~\ref{sec:visualization}, we illustrate the appropriate
visualization of NAME for facilitating its interpretation by a
non-expert and we discuss important aspects of the method

@@

ALE defines the local effect of the \(s\)-th feature on \(f(\cdot)\)
at point \((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
explanations at \(x_s\) are, then, weigthed by the conditional
distribution \(p(\xc|x_s)\) and are averaged, to produce the
summarized effect at \(x_s\):

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
%
The FE at \(x_s\) is the accumulation of the averaged
local effects from \(x_{s, min}\) until \(x_s\), i.e.
\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
\). As described at the Introduction, limiting the explanation to the
exepected value level does not shed light to possible heterogeneous
effects behind the averaged explanation. Therefore, we model the
uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
variance of the local explanations:

\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In
Section~\ref{sec:visualization}, we propose appropriate visualizations
for easier interpretation of Eq.~(\ref{eq:ALE_var}). In ALE, the
FE at \(x_s\) is the accumulation of the averaged local
effects from \(x_{min}\) until \(x_s\), as shown in
Eq.~(\ref{eq:ALE_accumulated_mean}). Equivalently, we define the
accumulated uncertainty (variance) until \(x_s\), as the integral of
the variances of the local effects:

\begin{equation}
  \label{eq:ALE_accumulated_var}
  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
\end{equation}
\noindent

The accumulated uncertainty is not a directly interpretable
quantity. It only helps us define a sensible objective for the
interval spliting step, as we discuss in
Section~\ref{sec:interval-spliting}.

@@

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\), so, the estimations are based
on the limited instances of the training set. Estimating
Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
point is impossible, because the probability of observing a sample
inside the interval \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We are, therefore, obliged to split the axis of \(x_s\)
into a sequence of non-overlaping intervals (bins) and estimate the
mean and the variance from the samples that lie inside each bin. The
mean effect at an interval \([z_1, z_2)\) is defined as the mean of
the expected effects:

\begin{equation}
  \label{eq:mu_bin}
  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
\end{equation}

\noindent
Similarly, the accumulated variance at an interval \([z_1, z_2)\) is
defined as:

\begin{equation}
  \label{eq:var_bin}
  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
\end{equation}
\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
at an interval \([z_1, z_2)\) is the accumulation of the all variances
plus the accumulation of squared residuals inside the interval:

\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is in the Appendix. Theorem 1 decouples the accumulated
variance at an interval into two terms. The first term
\(\int_{z_1}^{z_2} \sigma^2(z) \partial z\), quantifies the aggregated
uncertainty due to the natural characteristics of the experiment
\(\mathcal{H}\) and the second term adds extra nuisance uncertainty
due to the limited resolution \(\mathcal{H}_n\). In other words,
enforcing the computation of a single effect for all points in
\([z_1, z_2)]\), burdens the estimation inside the interval with a
nuisance uncertainty of:

\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \int_{z_1}^{z_2} \mathcal{H}(z) + \mathcal{H}_n(z) \partial z
\end{equation}

Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be directly estimated
from the set \(\mathcal{S}\) of the dataset instances with the
\(s\)-th feature lying inside the interval, i.e.,
\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \). The mean
effect at the interval, Eq.~(\ref{eq:mu_bin}) is approximated by:

\begin{equation}
  \label{eq:mean_estimation}
  \hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ]
\end{equation}

and the accumulated variance, Eq.~(\ref{eq:var_bin}) can be
approximated by

\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}^2(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}(z_1, z_2) \right )^2
\end{equation}

The approximation is unbiased only if the points are uniformly
distributed in \([z_1, z_2]\). Elaborate.

\subsection{Interval Spliting As A Clustering Problem}
\label{sec:interval-spliting}

% Instead, we propose treating the axis-spliting step as an unsupervised
% clustering problem. The objective of the clustering problem should
% fulfill in the best way to contradictory ojectives. First, secure
% robust estimations of the expected effect and the uncertainty inside
% each bin given the limited instances of the training set and, second,
% create bins with as homogeneous local effects as possible, for not
% losing fine-grain resolution FEs due to wide bins.

We formulate the axis-splitting as an unsupervised clustering
problem. We search for the optimal bin splitting, i.e., the number and
size of consecutive non-overlapping intervals that minimizes the
accumulated variance. The optimization problem is defined as follows:

\begin{equation}
  \label{eq:opt}
\begin{aligned}
\min_{ \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \\
\textrm{s.t.} \quad & |\mathcal{S}_k| \geq N\\
\end{aligned}
\end{equation}

The objective of the optimization problem is the sum of the
accumulated variances in each bin, estimated by the instances of the
training set. As we prove in Theorem 1,
\(\mathcal{L} = \sum_{k=1}^K \mathcal{H}_{bin}(z_{k-1}, z_k)\) is the
sum of the uncertainty of the local explanations \(\mathcal{H}\) due
to heterogeneous effects and the added nuisance uncertainty due to
bin-splitting \(\mathcal{H}_n\). Given that \(\mathcal{H}\) is not
affected by interval splitting, optimizing \(\mathcal{L}\) equals to
finding the sequence of non-overlaping bins that add the minimum
nuisance uncertainty. The restriction of (\ref{eq:opt}) secures that
each bin is populated with enough samples for a robust estimation,
which is set by the parameter \(N\). In other words, the user makes
the following proposal; \textit{find the bin sequence that adds the
  minimum nuisance uncertainty, given that I want at least \(N\)
  points per bin}.

\subsubsection{Algorithm For Solving The Clustering Problem}
\label{sec:dynamic-programing}

Solving the problem of finding (a) the optimal number of bins \(K\)
and (b) the optimal bin limits for each bin
\([z_{k-1}, z_k] \forall k\) to minimize:

\begin{equation}
  \label{eq:1}
  \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
\end{equation}
%
The constraints are that all bins must include more than \(\tau\)
points, i.e., \(|\mathcal{S}_k| \geq \tau\).

>>>>

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML system with the capability to
explain their predictions, apart from predicting with accuracy. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local ones into a single interpretable outcome, usually a
number or a plot. If a user wants to get a rough overview about which
features are significant (feature importance) or whether a particular
feature has a positive or negative effect on the output (feature
effect), they should opt for a global explainability technique. On
the other hand, aggregating the individual explanations for producing
a concise global one is vulnerable to misinterpretations; Under strong
feature interactions, the global explanation may obfuscate
heterogeneous effects~\citep{Herbinger2022repid} that exist under the
hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
of global explainability methods. The objective of FE is to the
isolate and visualize the impact of a single feature on the
output.~\footnote{FE methods also isolate the effect of a pair of
  features to the output. Combinations of more than two features are
  not usual, because they encounter, among others, visualization
  difficulties.} FE methods suffer from aggregation bias because,
often, the rationale behind the average effect might be unclear. For
example, a feature with zero average effect may indicate that the
feature has no effect on the output or, contrarily, it has a highly
positive effect in some cases and a highly negative in others. There
are three widely-used FE methods; Partial Dependence Plots
(PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML
problems. Therefore, ALE has been established as the state-of-the-art
FE method.

However, ALE faces two crucial drawbacks. First, it does not inform
the user about potential heterogeneous effects hidden behind the
average effect. In contrast, in the case of PDP, the heterogeneous
effects can be spotted by exploring the Individual Conditional
Expectations (ICE)\citep{goldstein2015peeking}. Second, ALE
approximation from the samples of the training set requires
partitioning the axis of the feature of interest in a sequence of
non-overlapping intervals and estimating a single effect from the
population of samples in each interval. We refer to this step using
the term \textit{bin-splitting problem}. Deciding on the sequence of
intervals has important implications because ALE's interpretation in
cases of correlated features is meaningful only inside each interval
\citep{molnar2022}. However, this crucial step has not been given the
appropriate attention, making the approximation vulnerable to
potential misinterpretations.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the explanation. The uncertainty of the
global explanation expresses how certain we are that the averaged
explanation is valid if applied to an instance drawn at random and
reveals the heterogeneous effects. Given that ALE's interpretation
(expected effect and uncertainty) is meaningful only inside each
interval, we design a principled framework, where we treat the
bin-splitting step as a data-driven clustering problem, searching for
the optimal splitting given available instances of the training
set. We, also, present a computationally-grounded algorithm for
finding the optimal solution.

\paragraph{Contributions.} The contribution of this paper is NAME, a
feature effect method that:

\begin{itemize}
\item Extends ALE with a probabilistic component to quantify the
  heterogeneous effects behind the global explanation.
\item Automatically extracts regions with similar effects, improving
  the estimation and the interpretability of ALE plots.
\end{itemize}

We provide empirical evaluation of the method in artificial and real
datasets. The implementation of our method and the code for
reproducing all the experiments is provided in the submission and will
become publicly available upon acceptance.

@@

\paragraph{Feature Effect Methods.} The three well-known feature
effect methods are: PDP, MP and ALE. PDP formulates the FE
of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects:
\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
ALE has specific advantages which gain particular value in cases of
correlated input features. In these cases, PDP integrates over
unrealistic instances, due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore a trust able FE method
in cases of correlated features.

@@

FE methods answer the question \textit{what is expected to happen to
  the output (expected effect), if the value of a specific feature is
  increased/decreased}. Having an answer to the question above, it
comes naturally to also wonder \textit{how certain we are about the
  expected change (uncertainty)}. For this reason, the quantification
of the uncertainty along with the expected effect has attracted a lot
of interest. The level of uncertainty is mostly quantified by
measuring the existence of heterogeneous effects, i.e. whether there
are local explanations that deviate from the expected global
effect. ICE and d-ICE plots provide a visual understanding of the
heterogeneous effects on top of PDPs. Another approach targets on
grouping the heterogeneous effects, e.g., allocating ICE plots in
homogeneous clusters, by dividing the input
space.\citep{molnar2020model} Some other approaches, like H-Statistic,
Greenwel, move a step behind and try to quantify the level of
interaction between the input features, a possible cause of
heterogeneous effects. In this case, the interpretation is indirect,
since a strong interaction index is only an indicator of heterogeneous
effects. The aforementioned approaches are under two limitations; They
either do not quantify the uncertainty of the FE directly or they are
based on PDPs, and, therefore, they are subject to the failure modes
of PDPs in cases of correlated features. To the best of our knowledge,
no method so far targets on quantifying the uncertainty of ALE.

@@

In real ML scenarios, the FE is estimated from the limited instances
of the training set.  \citep{apley2020visualizing} proposed dividing
the \(s\)-th axis in \(K\) bins and estimating the local effects in
each bin by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE) for
computing the local effects on the training instances using auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost. However, none of the approximations above deals
with the crucial problem of bin-splitting. As indicated
by~\citep{molnar2022}, in ALE the effects are computed per interval
(region) and the interpretation of the effect can only be
local.

@@

The NAME method extends ALE for measuring the uncertainty of the
expected effect and automatically extracts regions with homogeneous
effects. In Section~\ref{sec:NAME-definition} we define the component
for the uncertainty quantification. In
Section~\ref{sec:interval-based-estimation}, we show how to estimate
the average effect and the uncertainty on a specific interval from the
limited samples of the training set. We, also, make an important proof
about the aggregated variance defined over a bin. In
Section~\ref{sec:interval-spliting}, we define and solve the problem
of optimal bin-splitting for automatically extract regions with
homogeneous effects. Finally, in Section~\ref{sec:visualization}, we
illustrate the appropriate visualization of NAME for facilitating its
interpretation by a non-expert and we discuss important aspects of the
method.

@@

ALE defines the local effect of the \(s\)-th feature at point
\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local explanations at
\(x_s\) are, then, weighted by the conditional distribution
\(p(\xc|x_s)\) and are averaged, to produce the summarized effect at
\(x_s\):
\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
%
The FE at \(x_s\) is the accumulation of the averaged local effects
from \(x_{s, min}\)\footnote{\(x_{s,min}\) is the minimum value the
  \(s\)-th attribute can take. In estimations, it is the minimum
  \(s\)-th attribute value among the training set instances.} until
\(x_s\), i.e.
\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
\). Limiting the explanation to the expected value level does not
shed light to potential heterogeneous effects behind the averaged
explanation. Therefore, we model the uncertainty of the local effects
\(\mathcal{H}(x_s)\) with the variance of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In
Section~\ref{sec:visualization}, we propose appropriate visualizations
for easier interpretation of Eq.~(\ref{eq:ALE_var}). ALE defines the
FE at \(x_s\) as the accumulation of the averaged local effects from
\(x_{min}\) until \(x_s\). Equivalently, we define the accumulated
uncertainty (variance) until \(x_s\), as the integral of the variances
of the local effects:
\begin{equation}
  \label{eq:ALE_accumulated_var}
  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
\end{equation}
\noindent
The accumulated uncertainty is not a directly interpretable quantity,
since it measures the total uncertainty that characterizes the
\(s\)-th feature effect plot. As stated in the Introduction, ALE
outcomes should be interpreted at a local level, therefore, the only
\(\mathcal{H}(x_s)\) is the directly interpretable quantity. However,
the accumulated uncertainty will help us define a sensible objective
for the bin-splitting step, as we will discuss in
Section~\ref{sec:interval-spliting}.

@@

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\), so, the estimations are based
on the limited instances of the training set. Estimating
Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
point is impossible, because the probability of observing a sample
inside the interval \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We are, therefore, obliged to split the axis of \(x_s\)
into a sequence of non-overlapping intervals (bins) and estimate the
mean and the variance from the samples that lie inside each bin. The
mean effect at an interval \([z_1, z_2)\) is defined as the mean of
the expected effects:
\begin{equation}
  \label{eq:mu_bin}
  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
\end{equation}
\noindent
Similarly, the accumulated variance at an interval \([z_1, z_2)\) is
defined as:
\begin{equation}
  \label{eq:var_bin}
  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
\end{equation}
\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is in the Appendix. Theorem 1 decouples the total
uncertainty inside a bin
\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\) into two terms. The first term quantifies
the accumulated uncertainty due to the natural characteristics of the
experiment, i.e.,
\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}(z) \partial z\),
and the second term adds extra nuisance uncertainty due to limiting
the resolution
\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
\). In other words, the creation of a bin burdens the estimation with
a nuisance uncertainty:
\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
\end{equation}
Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be directly estimated
from the set \(\mathcal{S}\) of the dataset instances with the
\(s\)-th feature lying inside the interval, i.e.,
\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \). The mean
effect at the interval, Eq.~(\ref{eq:mu_bin}) is approximated by:
\begin{equation}
  \label{eq:mean_estimation}
  \hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ]
\end{equation}

and the accumulated variance, Eq.~(\ref{eq:var_bin}) can be
approximated by
\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}^2(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
The approximation is unbiased only if the points are uniformly
distributed in \([z_1, z_2]\). Elaborate.

\subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
\label{sec:interval-spliting}

As we discussed in the Introduction, when the features are (strongly)
correlated, ALE provides meaningful interpretations only inside a
bin. The accumulation of the mean effects inside each bin is done only
for a producing smooth curve for easier visual exploration. Therefore,
the extraction of regions with common effects is particularly
important for a correct interpretation.

We formulate the bin-splitting as an unsupervised clustering
problem. We want to partition the axis of the \(s\)-th feature into a
sequence of variable-size non-overlapping intervals, so that the
regions that will be created to provide meaningful explanations. A
region with a meaningful explanation means that it clusters together
points with similar expected (averaged) effects. On the other hand,
since we estimate all the quantities from the limited samples of the
training set, we need enough samples inside each bin for robustly
estimate the expected effect and the variance. Therefore, we transform
the bin-splitting problem to the following optimization problem:
\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) covering
the axis of the \(s\)-th feature, \(z_0 = x_{s,min}, z_K = x_{s,max}\)
that minimize the sum of bin variances. The constraint of at least
\(N\) points in each bin sets the limit for the minimum number of
points inside each bin, to robustly estimate the mean effect and the
variance. There is no constraint about the bin size and the number of
bins \(K\).

The objective function is the sum of the approximations of the bin
variances. As we prove in Theorem 1, the observable bin variance can
can be split in the accumulation of point-wise variances plus the
residuals. In Corollary 1, we show that the accumulation of the
point-wise variances is independent of the bin-splitting
procedure. Therefore by minimizing \(L\), we minimize the accumulation
of the point-wise residuals in each bin, searching for bins that
cluster together points with a common expected effect. An additional
interpretation is that we search for the bin-splitting that adds the
minimum the nuisance uncertainty over the lower-bound of the
uncertainty due to the natural characteristics of the experiment.

\subsubsection{Solve Bin-Splitting with Dynamic Programming}
\label{sec:dynamic-programing}
